Modeling memory and taxis to analyze migrations

### 2021-01-22 code/IndicesEquations.Rmd
refinements
Changed foraging efficiency function and added example of using indices for a model

$$O(t_1, t_2) = \int_D \sqrt{u(x, t_1) \, u(x, t_2)} \, dx; $$
After calcuating this overlap of the two matrices that contain the population, we find the values of $t_1$ and $t_2$ that minimize $O$, corresponding to the periods of maximum distance.  Thus, the migratoriness index is:

$$MI = min(O(t_1, t_2))$$

```{r}
 findMinOverlap <- function(P, world = world){

    getOverlap <- function(i1, i2, P){
      sum(sqrt(P[i1,] * P[i2,])) * world$dx
    }

    Overlap <- matrix(NA, world$tau, world$tau)
    for(i in 1:(world$tau-1))
      for(j in (i+1):world$tau)
        Overlap[i,j] <- sum(sqrt(P[i,] * P[j,])) * world$dx

    minOverlap <-  min(Overlap, na.rm = TRUE)
    list(times = which(Overlap == minOverlap, arr.ind = TRUE),
         overlap = minOverlap)
  }

@@

## Foraging Success

To quantify the consumers’ ability to track the distribution of their resources over space and time, we use the continuous form of the Bhattacharya Coefficient (BC; Bhattacharyya 1943).  The BC quantifies similarity between two distributions; we take the mean of the BC over the equilibrium year. Thus: 
$$BC = {1\over T} \int_T\int_D \sqrt{u(x,t) \, h(x,t)} \, dx\,dt$$
Where the spatial integral is taken over the domain ($D$).

The Bhattacharyya coefficient determines the probability of the overlap between the two distributions. Thus $0 \ge BC(p,q) \ge 1$. A larger Bhattacharyya coefficient indicates a smaller Bhattacharyya distance to show that the two distributions are similar. A larger distance indicates that the population is farther away from the resource at the given year.

```{r}
require(bio3d)
  BhattacharyyaDCoefficient <- function(i1, i2){
    BC <- bhattacharyya(cov(i1), cov(i2)) #do we need a q value? 
    list(BC)
  } #does this function take the mean of the BC over the year? 
```

We used the `bhattacharyya` function in the `bio3d` package (Grant et al. 2006).

@@

$$\overline{\sigma} = {1\over \tau} \int \sigma_u(t) \,dt$$

@@

and $\mu_u$ is the mean of $u$ at time $t$.  If the population is uniformly distributed over the domain, the standard deviation of the model would be $\frac{X_{max}}{\sqrt{12}}$.  Where therefore define the social cohesion index as: $$SC = 1-\frac{\sqrt{12} \, }{X_{max}}\overline{\sigma}$$

where $\sigma_{max}=\frac{1}{\sqrt{12}}.$  In theory, SC can be lower than 1 if, for example, the population concentrates in two groups at the edges of the domain, but in practice the SC is positive.  The maximum value of 1 would correspond to the entire population . 
```{r}
 getSD <- function(f, world){
    EX <- sum( world$X * f) * world$dx
    EX2 <- sum( world$X^2 * f) * world$dx
    sd  = sqrt(EX2 - EX^2)
  }

  computeCohesiveness <- function(P){
    SDs <- apply(P, 1, getSD, world = world)
    SD.max <- (1/sqrt(12))*max(world$X)
    SC <- 1 - SDs/SD.max
    c(SC.mean = mean(SC), SC.sd = sd(SC))
  }

@@

## References

Bhattacharyya, A. On a measure of divergence between two statistical populations defined by their probability distributions. Bull. Calcutta Math. Soc. 35 (1943), 99–109.

Grant, B.J. et al. (2006) Bioinformatics 22, 2695--2696.

>>>>

$$O(t_1, t_2) = \int_D \sqrt{u(x, t_1) \, u(x, t_2)} \, dx; $$
After calcuating this overlap of the two matrices that contain the population, we find the values of $t_1$ and $t_2$ that minimize $O$, corresponding to the periods of maximum distance. We subtract this minimum value to calculate migratoriness of the population. Thus, the migratoriness index (MI) is:

$$MI = 1-min(O(t_1, t_2))$$

```{r}
 computeMigratoriness <- function(pop, world){

  dx <- world$dx
  tau <- world$tau

  Overlap <- matrix(NA, tau, tau)
  for(i in 1:(tau-1))
    for(j in (i+1):tau)
      Overlap[i,j] <- sum(sqrt(pop[i,] * pop[j,])) * dx

  minOverlap <-  1-min(Overlap, na.rm = TRUE)
  list(times = which(Overlap == minOverlap, arr.ind = TRUE),
       overlap = minOverlap)
}

@@

## Foraging Efficiency

To quantify the consumers’ ability to track the distribution of their resources over space and time, we use the continuous form of the Bhattacharyya Coefficient (Bhattacharyya 1943).  The Bhattacharyya Coefficient quantifies similarity between two distributions; we take the mean of the coefficient over the equilibrium year to determine foraging efficiency (FE). Thus, the foraging efficiency index is: 
$$FE = {1\over \tau} \int_{0}^\tau \int_{0}^{X_{max}} \sqrt{u(x,t) \, h(x,t)} \, dx\,dt$$
Where the spatial integral is taken over the domain.

```{r}
computeEfficiency <- function(pop, resource, world){

  dx <- world$dx
  tau <- world$tau

  Similarity <- matrix(NA, tau, tau)
  for(i in 1:(tau-1))
    for(j in (i+1):tau)
      Similarity[i,j] <- sum(sqrt(pop[i,] * resource[j,])) * dx

  FS <- mean(Similarity, na.rm=TRUE)/tau
}
```

@@

$$\overline{\sigma} = {1\over \tau} \int \sigma_u(x,t) \,dt$$

@@

and $\mu_u$ is the mean of $u$ at time $t$.  If the population is uniformly distributed over the domain, the standard deviation of the model would be $\frac{X_{max}}{\sqrt{12}}$.  We therefore define the social cohesion (SC) index as: $$SC = 1-\frac{\sqrt{12} \, }{X_{max}}\overline{\sigma}$$

where $\sigma_{max}=\frac{1}{\sqrt{12}}.$  In theory, SC can be lower than 1 if, for example, the population concentrates in two groups at the edges of the domain, but in practice the SC is positive.  The maximum value of 1 would correspond to the entire population . 
```{r}
 getSD <- function(pop, world){
    EX <- sum( world$X * pop) * world$dx
    EX2 <- sum( world$X^2 * pop) * world$dx
    sd  = sqrt(EX2 - EX^2)
  }

  computeCohesiveness <- function(pop){
    SDs <- apply(pop, 1, getSD, world = world)
    SD.max <- (1/sqrt(12))*max(world$X)
    SC <- 1 - SDs/SD.max
    c(SC.mean = mean(SC), SC.sd = sd(SC))
  }

@@

## Combining the Indices
The following function calculates these three indices for a specific population and resource. 
```{r}
computeIndices <- function(pop, resource, world){
  SC = computeCohesiveness(pop, world)["SC.mean"]
  MI = computeMigratoriness(pop, world)$overlap
  FE = computeEfficiency(pop, resource, world)
  data.frame(SC, MI, FE)
}
```

## References  
Bhattacharyya, A. (1943). "On a measure of divergence between two statistical populations defined by their probability distributions". Bulletin of the Calcutta Mathematical Society. 35: 99–109. MR 0010358

### 2021-01-22 code/StrategicVsTactical_PartI_v2.Rmd
Changed foraging efficiency function and added example of using indices for a model

```{r, message=FALSE, warning=FALSE}
require(deSolve)
require(ReacTran)
require(magrittr)
```

@@

```{r}
source("functions2.R")
```

In the following example, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is fixed in the middle of the space (as above).  The memory is a one-year memory.  In this context, migration is not, in fact, optimal.  But the important piece is to see how the reponse mixes attraction to the resource with attraction to the previous year's behavior.  Note that the year here (for simplicity, speed and symmetry), is 100 days long.

@@

>>>>

```{r, message=FALSE, warning=FALSE}
source("functions2.R")
```

@@

In the following example, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is fixed in the middle of the space (as above).  The memory is a one-year memory.  In this context, migration is not, in fact, optimal.  But the important piece is to see how the response mixes attraction to the resource with attraction to the previous year's behavior.  Note that the year here (for simplicity, speed and symmetry), is 100 days long.

@@

## Example
We apply the three indices to an example model. In this model, we let the population be "initialized" via a sinusoidal migration pattern, while the resource is seasonal in a broad, long peak. The model uses the parameters $\epsilon = 5, \alpha = 300, \beta0=100, \beta1 = 300$ so that the migratory population combines both memory and a tactical strategy. We run the model for 12 years and compute the three indices. Then we plot the three indices to the view the trend of the migratory population over time.

```{r}
source("functions_indices.R")
  world$resource <- getPulsedResource(world$time, world$X, c(t.peak = 25, t.sd = 9, 
                                                             x.peak = 80, x.sd = 9))

  parameters <- c(epsilon = 5, alpha = 300, beta0=100, beta1 = 300)
  world.R1 <- world %>% 
    list_modify(resource = getPulsedResource(1:world$tau, world$X, 
                                             c(t.peak = 25, t.sd = 9, 
                                               x.peak = 80, x.sd = 9)))
  M <- runManyYears(world.R1, Parameters = parameters, n.years = 6)

  world.R1 <- world %>% list_modify(pop = M$Year6)
  M2 <- runManyYears(world.R1, Parameters = parameters, n.years = 6)

  source("functions_indices.R")
  indices1 <- sapply(M, computeIndices, resource = world$resource, world = world) %>% t
  indices2 <- sapply(M2, computeIndices, resource = world$resource, world = world) %>% t
  I <- rbind(indices1, indices2)

  matplot(I, type="o", pch = 19, lty = 1) #year 6 and year 7 are the same so it is plotted twice 
  legend("topright", pch = 19, col = 1:3, legend = colnames(I))
